---
title: "An introduction to Protein2Tissue"
author:
  - name: "Chaoran Yang"
    affiliation: "Victorian Heart Institute, Monash University"
  - name: "Yao Chen"
    affiliation: "Faculty of Science, Ghent University"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to Protein2Tissue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse  = TRUE,
    comment   = "#>",
    fig.width = 7,
    fig.height = 5
)
```

# Introduction

`Protein2Tissue` is a Bioconductor package for tissue enrichment analysis of
protein and gene lists. Given a set of proteins or genes of interest and a
background set, the package identifies which tissues are significantly
over-represented using statistical tests. Results can be explored through
a suite of visualisation functions including heatmaps, bar charts, network
plots, and GO enrichment analysis.

Gene expression reference data are derived from the
[Human Protein Atlas](https://www.proteinatlas.org) (HPA), which classifies
genes into three tissue-specific expression categories:

- **Tissue enriched** — expression is significantly higher in one tissue
  compared to all others
- **Group enriched** — elevated expression in a small group of tissues
- **Tissue enhanced** — expression is higher in one tissue relative to the
  average across all tissues
- **Not detect** - no expression detected in any tissue


## Installation

```{r install, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("ChrYang/Protein2Tissue")
```

Once accepted to Bioconductor, the package will also be installable via:

```{r install_bioc, eval = FALSE}
# not yet available - coming soon
BiocManager::install("Protein2Tissue")
```

## Loading the package

```{r load, message = FALSE, warning = FALSE}
library(tidyverse)
library(Protein2Tissue)
```

# Input data

## Reference database

The package ships with `ee_tb`, an example gene/protein tissue expression 
data frame derived from the Human Protein Atlas. It contains tissue-specific 
expression classifications for human genes. It can be inspected directly:

```{r database, eval = FALSE}
data("ee_tb", package = "Protein2Tissue")
head(ee_tb)
```

Key columns in `ee_tb`:

| Column | Description |
|---|---|
| `Gene` | HGNC gene symbol |
| `Ensembl` | Ensembl gene identifier |
| `tissue` | tissue name |
| `signal` | expression category |
| `Tissue.specificity` | RNA specificity class |
| `Is_secreted` | whether the protein is secreted |

# Scenario 1: Assessing Tissue-Specific Protein Expression

We are going to demonstrate how to test tissue enrichment of proteins using 
the example dataset within the `Protein2Tissue` package.


`Protein2Tissue` requires two inputs:

- **Input**: a list of proteins or genes of interest (for example, 
differentially expressed proteins from a plasma proteomics experiment).
- **Background**: the full set of proteins or genes having been 
detected in the experiment. If this is not provided, all genes or proteins 
in the provided database are used as the background.
```{r inputs}
# example input: gene symbols
data("plasma_data", package = "Protein2Tissue")
myList <- plasma_data %>%
  filter(Coefficient.Age > 0 & q.Age < 0.05) %>%
  pull(UniqueSymbol) %>%
  toupper()

# example background: Gene symbols of all detected proteins
myBackground <-  toupper(plasma_data$UniqueSymbol)
  
```

The identifier types used for the input and background do not have to match. 
Use `typeKey` and `typeKeyBg` to specify how each list should be matched to the 
tissue expression database.

## Step 1 — Tissue enrichment analysis

`tissueAnalysis()` maps your input and background lists against the reference 
database (here we are using the shipped dataset in the package derived from the
Human Protein Atlas database) and returns an `Enrichment` S4 object:

```{r tissueAnalysis, eval = FALSE}
result <- tissueAnalysis(
    input      = myList,
    background = myBackground,
    typeKey    = "Gene",      # identifier type for input
    typeKeyBg  = "Gene",   # identifier type for background
)

```

The `Enrichment` object has two slots:

```{r slots, eval = FALSE}
# matched input proteins with HPA annotations
result@input[1:6,1:6]

# matched background proteins with HPA annotations
result@background[1:6,1:6]
```

A custom database can be substituted if needed:

```{r custom_db, eval = FALSE}
result <- tissueAnalysis(
    input    = myList,
    database = myCustomDatabase
)
```

Enrichment results can be saved to disk as CSV or TSV files:

```{r custom_db, eval = FALSE}
writeOutput(result,type = "csv","~/Enrichment_Results.csv")
```

## Step 2 — RNA tissue specificity overview

Before running statistical tests, it is useful to inspect how your input
genes are distributed across tissue specificity classes:

```{r plotProteinType, eval = FALSE}
plotProteinType(result)
```

This bar chart shows how many of your input genes fall into each specificity 
class (tissue enriched, group enriched, tissue enhanced,
low specificity, etc.), giving a quick overview of the tissue-specific
character of your gene list.

## Step 3 — Expression fingerprint heatmap

`heatmapEnrich()` shows the tissue expression signal for each gene in your
input as a heatmap:

```{r heatmap, eval = FALSE}
# all signal categories
heatmapEnrich(result)

# restrict to strongly tissue-specific signals only
heatmapEnrich(
    result,
    inclusion = c("Tissue enriched", "Tissue enhanced")
)
```

Rows are genes and columns are tissues. Colour indicates the signal
category — orange cells indicate no tissue-specific signal for that
gene-tissue combination.

## Step 4 — Fisher's exact test

`fisherByTissue()` performs Fisher's exact test for each tissue to identify
which tissues are significantly enriched in your input list relative to the
background:

```{r fisher, eval = FALSE}
fisherResult <- fisherByTissue(
    result,
    inclusion     = "All",
    padj          = "BH",
    secretoryOnly = FALSE
)

head(fisherResult)
```

Each row in the output corresponds to one tissue and contains:

| Column | Description |
|---|---|
| `tissue` | tissue name |
| `input_in` | input proteins expressed in this tissue |
| `input_total` | total input proteins |
| `bg_in` | background proteins expressed in this tissue |
| `bg_total` | total background proteins |
| `input_prop` | proportion of input proteins in tissue |
| `bg_prop` | proportion of background proteins in tissue |
| `OR` | odds ratio |
| `p_value` | Fisher's exact test p-value |
| `p_adj` | adjusted p-value |

To restrict the analysis to secretory proteins only:

```{r fisher_secretory, eval = FALSE}
fisherResultSecretory <- fisherByTissue(
    result,
    padj          = "BH",
    secretoryOnly = TRUE
)
```

## Step 5 — Visualise Fisher results

`fisherPlot()` generates a horizontal grouped bar chart comparing input
and background proportions across tissues, annotated with p-values:

```{r fisherPlot, eval = FALSE}
fisherPlot(fisherResult)
```

Blue bars show the proportion of input proteins expressed in each tissue;
grey bars show the background proportion. P-values from Fisher's exact test
are annotated to the right of each tissue.

## Step 6 — Network visualisation

`networkPlot()` creates a bipartite network connecting genes to their
associated tissues, coloured by signal category:

```{r network, eval = FALSE}
networkPlot(result)
```

In the network:

- **Large circles** are tissue hubs, labelled with the tissue name
- **Small points** are genes
- **Edge colour** encodes the signal category (tissue enriched, group
  enriched, or tissue enhanced)

This is particularly useful for identifying groups of genes that share
tissue associations.

## Step 7 — GO enrichment analysis

`goEnriched()` performs Gene Ontology enrichment analysis on genes
showing tissue-specific expression in a tissue of interest:

```{r go, eval = FALSE}
# biological process enrichment for liver-associated genes
goResult <- goEnriched(
    result,
    type       = "BP",
    tissueTest = "Liver",
    inclusion  = "All"
)

# view top enriched terms
head(as.data.frame(goResult@result))

```

The `type` argument accepts:

- `"BP"` — Biological Process (default)
- `"MF"` — Molecular Function
- `"CC"` — Cellular Component

# Complete example workflow

```{r full_workflow, eval = FALSE}
library(Protein2Tissue)

# --- inputs ---
data("plasma_data", package = "Protein2Tissue")

myList <- plasma_data %>%
  filter(Coefficient.Age > 0 & q.Age < 0.05) %>%
  pull(UniqueSymbol) %>%
  toupper()

# example background: Ensembl IDs representing the detectable proteome
myBackground <-  toupper(plasma_data$UniqueSymbol)

# --- step 1: tissue enrichment ---
result <- tissueAnalysis(
    input      = myList,
    background = myBackground,
    typeKey    = "Gene",      # identifier type for input
    typeKeyBg  = "Gene",   # identifier type for background
)

writeOutput(result,type = "csv","~/Enrichment_Results.csv")

# --- step 2: overview ---
plotProteinType(result)

# --- step 3: heatmap ---
heatmapEnrich(result)

# --- step 4: Fisher's test ---
fisherResult <- fisherByTissue(result, padj = "BH")

# --- step 5: bar chart ---
fisherPlot(fisherResult)

# --- step 6: network ---
networkPlot(result)

# --- step 7: GO enrichment ---
goResult <- goEnriched(result, type = "BP", tissueTest = "Liver")
```

#Scenario 2: Tissue-Specific Expression Analysis of Differentially Expressed Genes
This package can also identify genes or proteins enriched in specified tissues 
from lists of differentially expressed genes (DEGs) derived from RNA‑seq data or
differentially expressed proteins (DEPs) obtained from tissue‑specific proteomics
experiments.

```{r full_workflow, eval = FALSE}


result2 <- tissueAnalysis2(
  input         = c("BRCA1", "EGFR", "ETDB", 
                    "CCDC7", "TSPY9", "CCDC148"), #List of DEG/DEPs
  typeKey       = "Gene",
  database      = NULL,      # NULL uses bundled HPA database
  inclusion     = "All",
  secretoryOnly = FALSE,
  tissueToTest  = "Testis"   # tissue name must match the database
)

```

The results can be written to disk by the `write.csv()`  function.

```{r full_workflow, eval = FALSE}
write.csv(result2,"~/result2.csv",row.names=F)
```

## Scenario 3: Customising Your Own Tissue Expression Database

In addition to using the bundled Human Protein Atlas–derived tissue expression 
database, this package allows you to construct a custom tissue expression 
database from a gene‑by‑tissue average expression matrix. This is useful when
working with user‑generated datasets, alternative expression atlases, or
organism‑specific expression resources. Help we are using the gene-by-tissue
average transcripts per million (TPM) matrix from GTEx bundled to this package 
as an example.

### Classifying tissue specificity from an expression matrix

The `geneClassification()` function classifies genes or proteins based on their 
tissue‑specific expression patterns. The input expression matrix should be 
formatted with genes (or proteins) as rows and tissues as columns.

```{r classify_genes_example, eval = FALSE}
# Example expression matrix (genes x tissues)
data("tissue_tpm_gtex")
# Classify tissue specificity using gene symbols
res <- geneClassification(
  expression          = tissue_tpm_gtex,
  typeKey             = "Ensembl",
  min_expression      = 1,
  secretory_analysis  = FALSE
)

head(res)

```


### Using predefined tissue groups
Related tissues can be grouped prior to classification, allowing tissue 
specificity to be assessed at the designated group rather than the automatically 
ranked group by expression levels. 

```r
# Define tissue groups
data("tissue_map")

# Classify genes using grouped tissues
res_grouped <- classify_genes(
  expression          = tissue_tpm_gtex,
  tissue_groups       = tissue_map,
  typeKey             = "Gene",
  min_expression      = 1,
  secretory_analysis  = FALSE
)
```

The resulting classification output can be used directly as a custom tissue 
expression database in downstream analyses, including tissue enrichment 
testing with `tissueAnalysis()`.


# Data source

The bundled tissue expression database uses tissue-specific gene 
expression data from the [Human Protein Atlas](https://www.proteinatlas.org). 
If you use `Protein2Tissue` in your work, please also cite the HPA:

> Uhlen M et al. (2015). Tissue-based map of the human proteome.
> *Science*, 347(6220). <https://doi.org/10.1126/science.1260419>

> Uhlen M et al. (2017). A pathology atlas of the human cancer
> transcriptome. *Science*, 357(6352).
> <https://doi.org/10.1126/science.aan2507>

# Session information

```{r sessionInfo}
sessionInfo()
```
